Supplementary material

### 2018-06-07 Manuscript/ms.tex
More cleaning

Our ability to effectively prevent the transmission of the dengue virus through targeted control of its vector, \emph{Aedes aegypti}, depends critically on our understanding of the link between mosquito abundance and human disease risk.
The mosquito and clinical surveillance data necessary to elucidate this link are widely collected, but have yet to be coupled with a modeling framework that accounts for the complex non-linear mechanisms involved in transmission, in particular the critical bottleneck imposed by mosquito mortality.
We developed a differential equation model of dengue transmission and embedded it in a Bayesian hierarchical framework that allowed us to estimate latent time series of mosquito demographic rates from mosquito trap counts and dengue case reports from the city of Vitoria, Brazil.
We used the fitted model to explore the effect of adult mosquito control on the disease burden in each year, while taking advantage of the Bayesian framework to carry through uncertainty from the posterior distributions of model parameters and mosquito demographic rates.
We found that control deployed between 30 and 40 weeks prior to the start of the year was most effective at preventing cases in that year. 
This timing generally coincided with a local minimum in the size of the susceptible pool.
Uncertainty in the effectiveness of control interventions was driven largely by uncertainty in the period of cross-immunity, with longer periods of cross-immunity leading to less effective control.
Understanding the dynamics of temporary human immunity is thus critical for understanding the effect of mosquito control efforts on human disease.

@@

Similarly, susceptible mosquitoes ($V_S$) become exposed ($V_E$) by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implement a gamma-distributed incubation period by chaining multiple exposed classes together (taking advantage of the fact that a gamma-distributed random variable can be generated through the sum of a collection of exponential random variables with the same rate parameter) \cite{Lloyd2001}.
Total mosquito population size ($V_N$) is controlled by a forced, seasonally varying growth rate ($r(t)$) and death rate ($d(t)$).
Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.

@@

The estimated weekly mosquito mortality rate varied seasonally, and was positively correlated with temperature \ref{mortality}.
The estimated oscillations in mosquito mortality were more complex than purely sinusoidal, with variations in shape from year to year.
The posterior mean of the baseline mortality rate, around which the time series oscillated, was 0.7/week, roughly half the prior mean of 1.47/week.
The posterior means of the $\epsilon_{\nu t}$ forcing the mosquito mortality process exhibited a higher-frequency periodic oscillation (Supplementary Figure 1), though the full posterior distribution of each $\epsilon_{\nu t}$ overlapped zero.
In addition, the standard deviation of the mortality forcing terms was small relative to the weak prior ($E(\sigma_{\nu}|\mathbf{y}) = 0.01$, Supplemental Figure 2).

@@

\begin{figure}[!h]
\includegraphics[angle = 270, scale = 0.8]{figures/fig5.eps}
\caption{{\bf The effect of cross-immunity on effect of control.}
The posteriod distribution of the maximum proportion of cases prevented by control in each year (left to right, 2009 to 2012), as a function of the period of cross-immunity ($1/\delta$). The shading indicates posterior density, with darkers shading indicating a larger density of posterior samples in that bin.
}
\label{immunity}
\end{figure}

>>>>

Our ability to effectively prevent the transmission of the dengue virus through targeted control of its vector, \emph{Aedes aegypti}, depends critically on our understanding of the link between mosquito abundance and human disease risk.
The mosquito and clinical surveillance data necessary to elucidate this link are widely collected, but have yet to be coupled with a modeling framework that accounts for the complex non-linear mechanisms involved in transmission.
Most critical are the bottleneck in transmission imposed by mosquito lifespan relative to the virus' incubation period, and the dynamics of human immunity.
We developed a differential equation model of dengue transmission and embedded it in a Bayesian hierarchical framework that allowed us to estimate latent time series of mosquito demographic rates from mosquito trap counts and dengue case reports from the city of Vitoria, Brazil.
We used the fitted model to explore the effect of adult mosquito control on the disease burden in each year, while taking advantage of the Bayesian framework to carry through uncertainty from the posterior distributions of model parameters and mosquito demographic rates.
We found that control deployed between 30 and 40 weeks prior to the start of the year was most effective at preventing cases in that year. 
This timing generally coincided with a local minimum in the size of the susceptible pool.
Uncertainty in the effectiveness of control interventions was driven largely by uncertainty in the period of cross-immunity, with longer periods of cross-immunity leading to less effective control.
Understanding the dynamics of temporary human immunity is thus critical for understanding the effect of mosquito control efforts on human disease.

@@

Similarly, susceptible mosquitoes ($V_S$) become exposed ($V_E$) by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implement a gamma-distributed incubation period by chaining multiple exposed classes together (taking advantage of the fact that a gamma-distributed random variable can be generated through the sum of a collection of exponential random variables with the same rate parameter) \cite{Lloyd2001}.
Total mosquito population size ($V_N$) is controlled by a forced, seasonally varying growth rate ($r(t)$), while the transmission bottleneck is captured with a forced, seasonally varying mortality rate ($d(t)$).
Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.

@@

The estimated weekly mosquito mortality rate varied seasonally, and was positively correlated with temperature (Figure \ref{mortality}).
The estimated oscillations in mosquito mortality were more complex than purely sinusoidal, with variations in shape from year to year.
The posterior mean of the baseline mortality rate, around which the time series oscillated, was 0.7/week, roughly half the prior mean of 1.47/week.
The posterior means of the $\epsilon_{\nu t}$ forcing the mosquito mortality process exhibited a higher-frequency periodic oscillation (Supplementary Figure 1), though the full posterior distribution of each $\epsilon_{\nu t}$ overlapped zero.
In addition, the standard deviation of the mortality forcing terms was small relative to the weak prior ($E(\sigma_{\nu}|\mathbf{y}) = 0.01$, Supplemental Figure 2).

@@

\begin{figure}[!h]
\includegraphics[angle = 270, scale = 0.8]{figures/fig5.eps}
\caption{{\bf The effect of cross-immunity on effect of control.}
The posterior distribution of the maximum proportion of cases prevented by control in each year (left to right, 2009 to 2012), as a function of the period of cross-immunity ($1/\delta$). The shading indicates posterior density, with darkers shading indicating a larger density of posterior samples in that bin.
}
\label{immunity}
\end{figure}

### 2018-06-07 Manuscript/supplement.tex
More cleaning

\begin{align}
y_t & \sim \text{NegBin}(\phi_y  (C_t - C_{t-1})N, \eta_y)\\
q_t & \sim \text{NegBin}((V_{Ct} - V_{Ct-1})N, \eta_q)\\
\\
\frac{dC}{dt} &= \rho E \\
\frac{dS}{dt} &= b - bS - \lambda V_{I}S + \delta R\\
\frac{dE}{dt} &= \lambda V_{I}S - (\rho + b)E\\
\frac{dI}{dt} &= \rho E - (\gamma + b)I\\
\frac{dR}{dt} &= \gamma I - (\delta + b)R\\
\\
\frac{dV_N}{dt} & = r V_N - \phi_q \tau_t V_N\\
\frac{dV_E}{dt} &= \lambda IV_S - (\rho_{vt} + d + \phi_q \tau_t)V_E\\
\frac{dV_I}{dt} &= \rho_{vt} V_E - (d + \phi_q \tau_t) V_I\\
\frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
V_S &= V_N - V_E - V_I\\
\\
d(t) &= d_0 \exp(\nu(t))\\
\frac{d^2\nu}{dt^2} &= -\omega^2 \nu + \epsilon_{\nu t}\\
\frac{d^2 r}{dt^2} &= -\omega^2 r + \epsilon_{rt},
\end{align}
where $\omega = 2\pi / 52$, and
\begin{align}
\epsilon_{\nu i} & \sim \text{Normal}(0, \sigma_{\nu})\\
\epsilon_{ri} & \sim \text{Normal}(0, \sigma_r),
\end{align}
for $i = 1 \dots 243$.

The following priors complete the model specification:
\begin{align}
\sigma_r & \sim \text{Half-normal}(0, 0.2)\\
\sigma_d & \sim \text{Half-normal}(0, 0.2)\\
-\log(\phi_q) & \sim \text{Half-normal}(13, 0.5)\\
\eta_y &\sim \text{Half-normal}(0, 5)\\
\eta_q & \sim \text{Half-normal}(0, 5)\\
T_{\rho} & \sim \text{Gamma}(8.3, 8.3)\\
\rho & = (0.87 T_{\rho})^{-1}\\
T_{\delta} & \sim \text{Gamma}(10, 10)\\
\delta & = (97 T_{\delta})^{-1}\\
\gamma' & \sim \text{Gamma}(100, 100)\\
\gamma & = 3.5\gamma'\\
\nu_0 & \sim \text{Normal}(0, 0.5)\\
r_0 & \sim \text{Normal}(0, 0.5)\\
\log(V_{N0}) & \sim \text{Normal}(0.7, 0.3) \\ 
S_0' & \sim \text{Beta}(4, 6)\\
E_0' & \sim \text{Gamma}(10, 0.1)\\
I_0' & \sim \text{Gamma}(6, 0.1)\\
E_0 & = E_0' / N \\
I_0 & = I_0' / N \\ 
S_0 & = S_0'(1 - E_0 - I_0) \\
R_0 & = 1 - S_0 - E_0 - I_0
\end{align}

@@

\item \textbf{Baselin mosquito mortality rate} ($d_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.

>>>>

\begin{align}
y_t & \sim \text{NegBin}(\phi_y  (C_t - C_{t-1})N, \eta_y)\\
q_t & \sim \text{NegBin}((V_{Ct} - V_{Ct-1})N, \eta_q)\\
\\
\frac{dC}{dt} &= \rho E \\
\frac{dS}{dt} &= b - bS - \lambda V_{I}S + \delta R\\
\frac{dE}{dt} &= \lambda V_{I}S - (\rho + b)E\\
\frac{dI}{dt} &= \rho E - (\gamma + b)I\\
\frac{dR}{dt} &= \gamma I - (\delta + b)R\\
\\
\frac{dV_N}{dt} & = r V_N - \phi_q \tau_t V_N\\
\frac{dV_{E1}}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E1}\\
\frac{dV_{E2}}{dt} &= \rho_{v}(t) V_{E1} - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E2}\\
\frac{dV_{E3}}{dt} &= \rho_{v}(t) V_{E2}  - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E3}\\
\frac{dV_{E4}}{dt} &= \rho_{v}(t) V_{E3}  - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E4}\\\frac{dV_I}{dt} &= \rho_{vt} V_{E4} - (d(t) + \phi_q \tau_t) V_I\\
\frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
V_S &= V_N - V_E - V_I\\
\\
d(t) &= d_0 \exp(\nu(t))\\
\frac{d^2\nu}{dt^2} &= -\omega^2 \nu + \epsilon_{\nu t}\\
\frac{d^2 r}{dt^2} &= -\omega^2 r + \epsilon_{rt},
\end{align}
where $\omega = 2\pi / 52$, and
\begin{align}
\epsilon_{\nu i} & \sim \text{Normal}(0, \sigma_{\nu})\\
\epsilon_{ri} & \sim \text{Normal}(0, \sigma_r),
\end{align}
for $i = 1 \dots 243$.

The following priors complete the model specification:
\begin{align}
\sigma_r & \sim \text{Half-normal}(0, 0.2)\\
\sigma_d & \sim \text{Half-normal}(0, 0.2)\\
-\log(\phi_q) & \sim \text{Half-normal}(13, 0.5)\\
\eta_y &\sim \text{Half-normal}(0, 5)\\
\eta_q & \sim \text{Half-normal}(0, 5)\\
T_{\rho} & \sim \text{Gamma}(8.3, 8.3)\\
\rho & = (0.87 T_{\rho})^{-1}\\
T_{\delta} & \sim \text{Gamma}(10, 10)\\
\delta & = (97 T_{\delta})^{-1}\\
\gamma' & \sim \text{Gamma}(100, 100)\\
\gamma & = 3.5\gamma'\\
d_0' & \sim \text{Gamma(10, 10)}\\
d_0 & = 1.47 d_0'\\
\nu_0 & \sim \text{Normal}(0, 0.5)\\
r_0 & \sim \text{Normal}(0, 0.5)\\
\log(V_{N0}) & \sim \text{Normal}(0.7, 0.3) \\ 
S_0' & \sim \text{Beta}(4, 6)\\
E_0' & \sim \text{Gamma}(10, 0.1)\\
I_0' & \sim \text{Gamma}(6, 0.1)\\
E_0 & = E_0' / N \\
I_0 & = I_0' / N \\ 
S_0 & = S_0'(1 - E_0 - I_0) \\
R_0 & = 1 - S_0 - E_0 - I_0
\end{align}

@@

\item \textbf{Baseline mosquito mortality rate} ($d_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
